---
title: "STAT 345 Midterm Project"
font: 12pt
date: "4/2/26"
output:
  word_document: default
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

```{r, echo = FALSE}
# Loading required packages
library(hoopR) # hoopR is used to get the basketball data
library(tidyverse)
library(ggthemes)
library(stringr)
library(ggplot2)
```

```{r, echo = FALSE}
# Loads play-by-play for all NBA teams in given seasons
pbp <- load_nba_pbp(seasons = 2018:2025)

# Filters only shooting plays by the Timberwolves
wolves_shots <- pbp %>%
  filter(shooting_play == TRUE, team_id == 16)

# Narrows the data set down to shot location, season occurred, and scoring play by using select so it easier to look at
# The scoring play will be the result of whether the shot was made or missed
wolves_shots_raw <- wolves_shots %>% 
  select(scoring_play, coordinate_x_raw, coordinate_y_raw, season)

# Adds a column to better show if a shot was made or missed
# Made or missed will be used important for the plots
wolves_shots_result <- wolves_shots_raw %>%
  mutate(result = factor(scoring_play,
                    levels = c(TRUE, FALSE),
                    labels = c("Made", "Missed")))

```

```{r, echo = FALSE}
# Drawing the NBA court
draw_court_scaled <- function(color = "black") { 
  
  # Function used to create circle that will be used for court drawing
  circle <- function(center = c(0, 0), r = 1, npoints = 100) {
    data.frame(x = center[1] + r * cos(seq(0, 2*pi, length.out = npoints)),
      y = center[2] + r * sin(seq(0, 2*pi, length.out = npoints)))
  }

# Defines angles for 3-point arc
theta = seq(acos(22/23), acos(-22/23), length.out = 200)

  # List of important NBA court markings to draw court
  list(
    # Outer boundary
    geom_rect(aes(xmin = 0, xmax = 50, ymin = 0, ymax = 48),
              fill = NA, color = color),

    # Hoop
    geom_path(data = circle(c(25, 5), 1.5), aes(x, y), color = color),

    # Backboard
    geom_segment(aes(x = 22, y = 4, xend = 28, yend = 4), color = color),
    
    # Restricted area
    geom_path(data = data.frame(x = 25 + 4 * cos(seq(0, pi, length.out = 100)),
    y = 5 + 4 * sin(seq(0, pi, length.out = 100))),
    aes(x, y), color = color),

    # Paint
    geom_rect(aes(xmin = 17, xmax = 33, ymin = 0, ymax = 19),
              fill = NA, color = color),

    # Free throw circle
    geom_path(data = circle(c(25, 19), 6), aes(x, y), color = color),

    # 3-point arc
    geom_path(data = data.frame(x = 25 + 23 * cos(theta), y = 5 + 23 * sin(theta)),
      aes(x, y), color = color),

    # Corner 3 lines
    geom_segment(aes(x = 3, y = 0, xend = 3, yend = 12), color = color),
    geom_segment(aes(x = 47, y = 0, xend = 47, yend = 12), color = color)
  )
}
# Combining all the knowledge together to plot the shots
ggplot(wolves_shots_result, aes(x = coordinate_x_raw, y = coordinate_y_raw + 4)) + 
  # Added 4 to y to shift shots to better fit court drawing as basket is considered 0 and anything below is negative 
  
  # NBA court that was created in above code
  draw_court_scaled() +
  
  # Reduces overlap from similar shot coordinates and improves visibility of shot density patterns
  geom_jitter(aes(color = result), width = .35, height = .35, alpha = 0.4, size = 1) +
  
  # Creates separate plots for each season
  facet_wrap(~season, nrow = 2) +
  
  
  # Customized colors to reflect team colors
  # Colors will reflect made or missed shot
  scale_color_manual(
    values = c("Missed" = "grey", "Made" = "navy"),
    labels = c("Made", "Missed")) +
  
  # Set limits to match court dimensions
  coord_fixed(xlim = c(0, 50), ylim = c(0, 47)) +
  
  # Adding title and labels
  labs(title = "Minnesota Timberwolves Shot Chart",
    subtitle = "Seasons 2018-2025", color = "Shot Result") +
  
  # Removes grid background
  theme_void()

```

## Executive Summary

-   Shot selection has shifted toward more perimeter shooting over time

-   Increased clustering of shots beyond the 3-point arc in recent
    seasons

-   Mid-range shot attempts have declined significantly

-   Shot distribution near the basket remains consistent and
    high-frequency

-   No major change in overall shot success patterns, but efficiency
    appears higher in modern shot zones

## Detailed Explanation

#### Increase in Perimeter Shooting

Gradually the Timberwolves are taking a larger portion of shots beyond
the 3-point arc. This follows the league wide trend emphasizing spacing and
higher-value shots.

#### Growth in 3-Point Shot Clusters

Recent seasons show greater denser shot cluster along the 3-point arc
and corner 3's. With shots beyond the arc being worth more points teams
are starting to shoot there more to maximize their points per
possession.

#### Decline in Mid-Range Shot Attempts

The mid-range shots have significantly declined over the years. This aligns with modern analytics on discouraging lower-efficiency shots. Mid-range is lower-efficiency because they are not as high percentage as close shots and worth as much as 3-point shots.

#### Consistent Rim Attacks

Shots in the paint and near the rim continue to be highly populated.
These are typically higher percentage looks giving teams a great
opportunity to maximize scoring efficiency.

#### Efficiency Patterns

With less mid-range shots, more 3-point, and a high frequency of close
shots teams are trending towards shooting the shots that give them the
best expected value and efficiency. The high success of close shots
keeps players shooting them and higher value of 3-pointers is what is
causing the shift to farther jump-shots.


## NBA 4-Point Line

If the NBA would add a 4-point arc it should be placed at 31 feet away
from the basket. At this distance players make about 27% of shots which
leads to a expected value of 1.08 which is the same as the league
3-point expected value of 1.08. The arc will be 31 feet away at all
spots and shouldn't have any corner areas. A positive to a 4-point arc
would be that it would create more spacing on the offense which could
lead to better driving lines and offensive creativity. A negative could
be the incentive to take low percentage shots and that could lead to
less made shots. With less made shots the NBA could lose viewers. In
2020, Stephen Curry made 41/101 from 30+ feet. If the 4-point shot was
added in this season he would have averaged 0.6 more points per game
bringing his average to 32.6.

```{r, echo = FALSE}
library(pracma)

# Find the hypotenuse of each shot given the x and y coordinates to find the distance from the hoop
wolves_shots_distance <- wolves_shots_result %>%
  mutate(shot_distance = hypot(x = coordinate_x_raw - 25, y = coordinate_y_raw))

# Groups shots into distance bins
# Shots are grouped into 1 foot bins to show the field percentage from each distance away from the hoop.
distance_summary <- wolves_shots_distance %>%
  mutate(distance_bin = cut(shot_distance, breaks = seq(0, 40, by = 1))
  ) %>%
  group_by(distance_bin) %>%
  summarise(
    attempts = n(), makes = sum(result == "Made"), fg_pct = makes / attempts)

# Calculating the expected value of shots
# Expected value = probability of making shot * point value
# This allows us to compare efficiency of 3-point vs potential 4-point shots
distance_summary <- distance_summary %>%
  mutate(
    expected_points_3 = fg_pct * 3,
    expected_points_4 = fg_pct * 4)
```
